---
title: "Differential-expression designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differential-expression designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`rsemflow differential deseq2` fits one DESeq2 model and writes one table per
comparison. Three options define the analysis:

| Option | Meaning |
| --- | --- |
| `--p-formula` | the full model, e.g. `'~ batch + condition'` |
| `--p-term` | which term's results to write |
| `--p-reference COLUMN::LEVEL` | the reference (denominator) level of a factor; repeatable |

Quote the formula so the shell does not split it. Each section below shows the
metadata, the command, the equivalent R call, and the tables produced.

## Two groups

```text
sample_id	condition
C1	Control
C2	Control
C3	Control
D1	Drug
D2	Drug
D3	Drug
```

```bash
rsemflow differential deseq2 --i-study study/ \
  --p-formula '~ condition' --p-term condition \
  --p-reference condition::Control \
  --o-differential results/differential/
```

```{r}
differential_deseq2(study, ~ condition, "condition", reference = "condition::Control")
```

Output: `condition_Drug_vs_Control.tsv`. Positive `log2FoldChange` means higher
in Drug.

## Adjusting for batch

Put the nuisance variable before the variable of interest:

```bash
--p-formula '~ batch + condition' --p-term condition --p-reference condition::Control
```

Batch and condition must not be confounded. If every Control sample was in batch
A and every Drug sample in batch B, the two effects cannot be separated and
rsemflow stops with "The design matrix is not full rank".

## Paired samples

```text
sample_id	patient	condition
P01_pre	P01	Pre
P01_post	P01	Post
P02_pre	P02	Pre
P02_post	P02	Post
```

```bash
--p-formula '~ patient + condition' --p-term condition --p-reference condition::Pre
```

Including `patient` compares Post with Pre within each patient. Output:
`condition_Post_vs_Pre.tsv`.

## More than two groups

```bash
--p-formula '~ treatment' --p-term treatment --p-reference treatment::Vehicle
```

Every level is compared with the reference: `treatment_DrugA_vs_Vehicle.tsv`,
`treatment_DrugB_vs_Vehicle.tsv`. Add other pairs with `--p-contrast
COLUMN::NUMERATOR::DENOMINATOR` (repeatable):

```bash
--p-contrast treatment::DrugB::DrugA
```

```{r}
differential_deseq2(
  study, ~ treatment, "treatment",
  reference = "treatment::Vehicle",
  contrast = "treatment::DrugB::DrugA"
)
```

## A numeric covariate

When the term is numeric, the table reports the change in log2 expression per
unit of the variable:

```bash
--p-formula '~ sex + age + response_score' --p-term response_score
```

Output: `response_score.tsv`, with `contrast = per_unit`. Columns are numeric
when every value parses as a number; force a type at import with
`--p-categorical` or `--p-numeric`. For example, sampling times 0, 6, and 24 h
analysed as three groups:

```bash
rsemflow data import --i-rsem-dir data/ --m-metadata metadata.tsv \
  --p-categorical time --o-study study/
```

## Interaction

Does the drug response differ between genotypes?

```bash
--p-formula '~ genotype * treatment' \
--p-term genotype:treatment \
--p-reference genotype::WT --p-reference treatment::Vehicle
```

One table per interaction coefficient, named after the DESeq2 coefficient
(`genotypeKO.treatmentDrug.tsv`). Its `log2FoldChange` is the drug effect in KO
minus the drug effect in WT.

In an interaction model, `--p-term treatment` on its own gives the drug effect
**at the reference genotype (WT)**. rsemflow warns about this.

## The effect within one group (simple effect)

To get the drug effect within KO from the same interaction model, add one
`--p-contrast` and one `--p-at`:

```bash
--p-formula '~ genotype * treatment' \
--p-term treatment \
--p-reference genotype::WT --p-reference treatment::Vehicle \
--p-contrast treatment::Drug::Vehicle \
--p-at genotype::KO
```

```{r}
differential_deseq2(
  study, ~ genotype * treatment, "treatment",
  reference = c("genotype::WT", "treatment::Vehicle"),
  contrast = "treatment::Drug::Vehicle",
  at = "genotype::KO"
)
```

Output: `treatment_Drug_vs_Vehicle_at_genotype_KO.tsv`, computed as the main
effect plus the KO interaction coefficient (recorded in `contrasts.tsv`). The
package tests check that this equals the drug effect obtained by refitting with
KO as the reference genotype.

In version 0.1.0, `--p-at` accepts exactly one `--p-at` and one `--p-contrast`,
and the contrast denominator must be the reference level.

## Any effect of time (likelihood-ratio test)

To ask whether expression changes over time at all, compare a full and a
reduced model:

```bash
--p-formula '~ patient + time' \
--p-test lrt --p-reduced '~ patient' \
--p-term time
```

```{r}
differential_deseq2(
  study, ~ patient + time, "time",
  test = "lrt", reduced = ~ patient
)
```

Output: `time_omnibus.tsv`. The reduced formula must be nested in the full one
and must drop the term. Because the test has no single direction,
`log2FoldChange` and `lfcSE` are `NA`, and the table cannot be used for
preranked GSEA.

## What every run records

| File | Content |
| --- | --- |
| `design-matrix.tsv` | the model matrix actually fitted, one row per sample |
| `contrasts.tsv` | one row per result table: name, term, contrast, DESeq2 coefficient |
| `analysis-info.tsv` | formula, term, test, reduced formula, alpha |

## What rsemflow checks, and what it does not

Checked before fitting: model variables exist and have no missing values, the
term is part of the formula, reference and contrast levels exist, and the design
matrix is full rank.

Not checked: whether you have enough biological replicates, whether the model
is appropriate for your experiment, or whether samples should be excluded.
Log fold changes are not shrunk.
