---
title: "Installation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Requirements

- R 4.3 or newer.
- Bioconductor packages: tximport, DESeq2, SummarizedExperiment,
  AnnotationDbi, fgsea, GSVA, and the annotation packages org.Hs.eg.db and
  org.Mm.eg.db.
- CRAN packages: data.table and msigdbr.

rsemflow supports human and mouse data.

## Option 1: install the R package from GitHub

Use this if you work in R, or on Windows without WSL.

```{r}
install.packages("BiocManager")
BiocManager::install("DaikiKumakura/rsemflow")
```

`BiocManager` installs the CRAN and Bioconductor dependencies from the right
repositories. The OrgDb annotation packages are optional dependencies; install
them if you will use `annotation ensembl`:

```{r}
BiocManager::install(c("org.Hs.eg.db", "org.Mm.eg.db"))
```

## Option 2: install from source with the command-line launcher

On Linux, macOS, WSL, or an HPC login node, the installer also puts an
`rsemflow` command on your `PATH`:

```bash
git clone https://github.com/DaikiKumakura/rsemflow.git
cd rsemflow
bash install_local.sh
```

The installer

1. checks that `R` and `Rscript` are available;
2. installs missing dependencies (`scripts/install_dependencies.R`);
3. runs `R CMD INSTALL .`;
4. links the launcher into `$HOME/.local/bin` (or `$RSEMFLOW_BIN`);
5. runs `rsemflow --version` to confirm.

Options, set as environment variables:

| Variable | Effect |
| --- | --- |
| `RSEMFLOW_BIN=/path/bin` | install the launcher somewhere else |
| `RSEMFLOW_SKIP_DEPS=1` | skip the dependency step (e.g. dependencies are managed by your cluster) |

If `$HOME/.local/bin` is not on your `PATH`, add it (for example in
`~/.bashrc`):

```bash
export PATH="$HOME/.local/bin:$PATH"
```

## Running the command line on Windows

Without WSL, call the same command-line interface through `Rscript`; everything
after `-e "rsemflow::cli_main()"` is passed to rsemflow:

```bash
Rscript -e "rsemflow::cli_main()" data import --i-rsem-dir data --m-metadata metadata.tsv --o-study study
```

## Check the installation

```bash
rsemflow --version
rsemflow --help
```

Then run the bundled example, which needs no annotation database or download:

```bash
bash inst/extdata/example/run_example.sh example-output
```

or, from R:

```{r}
library(rsemflow)
ex <- system.file("extdata", "example", package = "rsemflow")
study <- read_rsem_study(file.path(ex, "data"), file.path(ex, "metadata.tsv"))
study
```

## MSigDB gene sets

`enrichment gsea` and `enrichment gsva` use MSigDB through the msigdbr package
unless you supply your own GMT file. Depending on your msigdbr version, the
first use of a collection may download and cache data, so run it once on a
machine with internet access. Custom GMT files (`--i-genesets`) need no
download.

## Troubleshooting

| Message | Cause and fix |
| --- | --- |
| `rsemflow: command not found` | The launcher directory is not on `PATH`; see above. |
| `R and Rscript must be on PATH` | Install R, or add its `bin` directory to `PATH`. |
| `Annotation package 'org.Hs.eg.db' is not installed` | `BiocManager::install("org.Hs.eg.db")` (or `org.Mm.eg.db`). |
| `Study format version ... is not supported` | The study was written by a different rsemflow version; run `rsemflow data import` again. |
