---
title: "Get started with rsemflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with rsemflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

rsemflow takes the `*.genes.results` files written by RSEM and a sample
metadata table, and produces the standard downstream results of a bulk RNA-seq
experiment as tab-separated tables:

- expression summaries per sample and per gene,
- normalized and variance-stabilized expression,
- PCA and sample correlation,
- DESeq2 differential expression for a design you specify,
- preranked GSEA and GSVA/ssGSEA pathway scores,
- gene annotation.

It does not draw figures. Every table can be opened in R, Python, Excel, or
Prism.

## What you need

```text
project/
├── data/
│   ├── S01.genes.results
│   ├── S02.genes.results
│   └── ...
└── metadata.tsv
```

`metadata.tsv` is an ordinary table with a `sample_id` column that matches the
file names (`S01.genes.results` → `S01`), plus any columns that describe the
samples:

```text
sample_id	condition	batch
S01	Control	A
S02	Control	B
S03	Drug	A
S04	Drug	B
```

## Three commands

Once installed (see `vignette("installation", package = "rsemflow")`), a first
analysis is three commands:

```bash
rsemflow data import --i-rsem-dir data/ --m-metadata metadata.tsv --o-study study/
rsemflow expression pca --i-study study/ --o-pca results/pca/
rsemflow differential deseq2 \
  --i-study study/ \
  --p-formula '~ batch + condition' \
  --p-term condition \
  --p-reference condition::Control \
  --o-differential results/differential/
```

`data import` checks the input and stores it as a *study* directory. Every
later command reads that study, so the RSEM files are parsed and validated once.

## The same analysis in R

Each command is an ordinary R function. Here they run on the synthetic example
shipped with the package (6 samples, 200 genes):

```{r}
library(rsemflow)

ex <- system.file("extdata", "example", package = "rsemflow")
study <- read_rsem_study(
  rsem_dir = file.path(ex, "data"),
  metadata = file.path(ex, "metadata.tsv")
)
study
```

```{r}
pca <- compute_pca(study)
pca$variance[1:3, ]
```

```{r}
de <- differential_deseq2(
  study,
  formula = ~ batch + condition,
  term = "condition",
  reference = "condition::Control"
)
names(de$results)
head(de$results$condition_Drug_vs_Control[, c("gene_id", "baseMean", "log2FoldChange", "padj")])
```

The example data are synthetic, so the numbers carry no biological meaning.

## How rsemflow treats your design

rsemflow never chooses a model for you. The formula (`~ batch + condition`), the
term to report (`condition`), and the reference level (`Control`) are always
explicit. In exchange, it stops instead of returning results that cannot be
trusted: when sample IDs do not match the metadata, when gene tables differ
between files, when a model variable has missing values, or when the design is
confounded (for example, every Control sample in batch A and every Drug sample
in batch B).

## Next steps

- `vignette("installation", package = "rsemflow")`: installing the R package
  and the `rsemflow` command.
- `vignette("tutorial", package = "rsemflow")`: a full command-line walkthrough
  with every output file.
- `vignette("designs", package = "rsemflow")`: paired, multi-group, numeric,
  interaction, and time-course designs.
