---
title: "Tutorial: a complete command-line analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial: a complete command-line analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rsemflow)

# Copy the bundled example into a scratch directory and run every chunk there,
# so the commands below are exactly the ones a reader would type.
work <- file.path(tempdir(), "rsemflow-tutorial")
unlink(work, recursive = TRUE)
dir.create(work)
ex <- system.file("extdata", "example", package = "rsemflow")
file.copy(file.path(ex, "data"), work, recursive = TRUE)
file.copy(file.path(ex, "metadata.tsv"), work)
knitr::opts_knit$set(root.dir = work)

# Each `bash` block is shown as typed; the hidden chunk after it runs the same
# command through rsemflow::cli_main() and prints what the command prints.
rsemflow_cmd <- function(cmd) {
  args <- scan(text = cmd, what = "", quiet = TRUE)
  cli_main(args[-1])
}
```

This tutorial runs the whole command-line workflow on the synthetic example
that ships with rsemflow: six samples (three Control, three Drug, in two
batches) and 200 genes. Every output shown below was produced by the commands
on this page when the documentation was built.

The data are synthetic; the numbers illustrate the file formats, not biology.

## 0. The input

All outputs are plain TSV files. To look at them, this page uses a small
helper that reads a table and shows its first rows:

```{r}
show_table <- function(path, n = 6, cols = NULL) {
  x <- utils::read.delim(path, check.names = FALSE)
  if (!is.null(cols)) x <- x[, cols, drop = FALSE]
  utils::head(x, n)
}
```

```text
.
├── data/
│   ├── C1.genes.results   C2.genes.results   C3.genes.results
│   └── D1.genes.results   D2.genes.results   D3.genes.results
└── metadata.tsv
```

```{r}
show_table("metadata.tsv")
```

Each RSEM file is the standard gene-level output:

```{r}
show_table("data/C1.genes.results", n = 3)
```

## 1. Import

```bash
rsemflow data import --i-rsem-dir data/ --m-metadata metadata.tsv --o-study study/
```

```{r, echo = FALSE}
rsemflow_cmd("rsemflow data import --i-rsem-dir data/ --m-metadata metadata.tsv --o-study study/")
```

Import checks that every RSEM file has the required columns, that file names
and `sample_id` match one-to-one, and that all files share the same gene table.
It then reads the counts with tximport and writes a study directory:

```{r}
list.files("study")
show_table("study/study-info.tsv", n = 10)
```

`expected_count.tsv`, `tpm.tsv`, and `effective_length.tsv` are gene × sample
matrices. `study.rds` is what later commands read.

`age` contains only numbers, so it was treated as numeric; `condition` and
`batch` became categorical. If a numeric-looking column is really a set of
groups (for example time points 0, 6, 24), add `--p-categorical time` to
`data import`.

## 2. Expression summaries

```bash
rsemflow data summarize --i-study study/ --o-summary results/summary/
```

```{r, echo = FALSE}
rsemflow_cmd("rsemflow data summarize --i-study study/ --o-summary results/summary/")
```

```{r}
show_table("results/summary/samples.tsv")
show_table("results/summary/genes.tsv", n = 3)
```

These are expression-level summaries computed from the RSEM tables, not
alignment QC. rsemflow does not flag or remove samples; that decision stays
with you.

## 3. Sample structure: PCA and correlation

```bash
rsemflow expression pca --i-study study/ --o-pca results/pca/
```

```{r, echo = FALSE}
rsemflow_cmd("rsemflow expression pca --i-study study/ --o-pca results/pca/")
```

By default PCA uses the variance-stabilized counts (DESeq2 VST) of the 500 most
variable genes (here all 200), centred and not scaled.

```{r}
show_table("results/pca/scores.tsv", cols = c("sample_id", "PC1", "PC2", "PC3"))
show_table("results/pca/variance.tsv", n = 3)
```

`loadings.tsv` holds the gene loadings and `analysis-info.tsv` records the
transformation and number of genes used.

```bash
rsemflow expression correlation --i-study study/ --o-correlation results/correlation.tsv
```

```{r, echo = FALSE}
rsemflow_cmd("rsemflow expression correlation --i-study study/ --o-correlation results/correlation.tsv")
```

```{r}
corr <- show_table("results/correlation.tsv")
corr[, -1] <- round(corr[, -1], 3)
corr
```

Use `--p-method spearman` or `--p-transform log2-tpm` to change the defaults.

## 4. Normalized expression

```bash
rsemflow expression normalize --i-study study/ --p-method vst --o-expression results/vst.tsv
```

```{r, echo = FALSE}
rsemflow_cmd("rsemflow expression normalize --i-study study/ --p-method vst --o-expression results/vst.tsv")
```

`--p-method` accepts `vst`, `normalized-counts` (DESeq2 size factors), and
`log2-tpm` (`log2(TPM + 1)`).

## 5. Differential expression

The samples come from two batches, so batch goes into the model and the
Drug-versus-Control effect is estimated within batch:

```bash
rsemflow differential deseq2 \
  --i-study study/ \
  --p-formula '~ batch + condition' \
  --p-term condition \
  --p-reference condition::Control \
  --o-differential results/differential/
```

```{r, echo = FALSE}
rsemflow_cmd("rsemflow differential deseq2 --i-study study/ --p-formula '~ batch + condition' --p-term condition --p-reference condition::Control --o-differential results/differential/")
```

```{r}
list.files("results/differential")
de <- utils::read.delim("results/differential/condition_Drug_vs_Control.tsv")
de <- de[order(de$padj), ]
utils::head(de[, c("gene_id", "contrast", "baseMean", "log2FoldChange", "lfcSE", "stat", "padj")])
```

The first six columns identify the comparison; the rest are the DESeq2 result
columns, with unshrunken log2 fold changes. The three other files record exactly
what was fitted:

```{r}
show_table("results/differential/analysis-info.tsv", n = 10)
show_table("results/differential/contrasts.tsv")
show_table("results/differential/design-matrix.tsv")
```

For paired, multi-group, numeric, interaction, and time-course designs, see
`vignette("designs", package = "rsemflow")`.

## 6. Preranked GSEA

GSEA ranks genes by the DESeq2 Wald statistic (`stat`). With real human or
mouse data, MSigDB collections are one option away:

```bash
rsemflow enrichment gsea \
  --i-differential results/differential/condition_Drug_vs_Control.tsv \
  --p-species human --p-collection hallmark \
  --o-enrichment results/gsea.tsv
```

The example genes are synthetic and match no MSigDB set, so here we use a
custom GMT file: genes 1–20, genes 21–40, and genes 101–150.

```{r}
ids <- sprintf("ENSG%011d", 1:200)
writeLines(c(
  paste(c("GENES_001_020", "synthetic", ids[1:20]), collapse = "\t"),
  paste(c("GENES_021_040", "synthetic", ids[21:40]), collapse = "\t"),
  paste(c("GENES_101_150", "synthetic", ids[101:150]), collapse = "\t")
), "sets.gmt")
```

```bash
rsemflow enrichment gsea \
  --i-differential results/differential/condition_Drug_vs_Control.tsv \
  --i-genesets sets.gmt \
  --p-min-size 10 \
  --o-enrichment results/gsea.tsv
```

```{r, echo = FALSE}
rsemflow_cmd("rsemflow enrichment gsea --i-differential results/differential/condition_Drug_vs_Control.tsv --i-genesets sets.gmt --p-min-size 10 --o-enrichment results/gsea.tsv")
```

```{r}
gsea <- utils::read.delim("results/gsea.tsv")
gsea$leadingEdge <- substr(gsea$leadingEdge, 1, 40)
gsea[, c("pathway", "size", "NES", "pval", "padj", "leadingEdge")]
```

`leadingEdge` lists genes separated by `;`. p-values are estimated by sampling,
so rsemflow fixes the seed (`--p-seed`, default 42) and repeated runs give the
same table.

## 7. GSVA and ssGSEA scores

```bash
rsemflow enrichment gsva \
  --i-study study/ \
  --i-genesets sets.gmt \
  --p-min-size 10 \
  --o-scores results/gsva.tsv
```

```{r, echo = FALSE}
rsemflow_cmd("rsemflow enrichment gsva --i-study study/ --i-genesets sets.gmt --p-min-size 10 --o-scores results/gsva.tsv")
```

```{r}
gsva <- utils::read.delim("results/gsva.tsv", check.names = FALSE)
gsva <- gsva[, -(2:4)]
gsva[, -1] <- round(gsva[, -1], 3)
gsva
```

Scores are computed per sample from `log2(TPM + 1)`; add `--p-method ssgsea`
for single-sample GSEA. rsemflow does not test scores between groups.

## 8. Gene annotation

Annotation adds gene symbols and biotypes while keeping the original RSEM
`gene_id`. The most faithful source is the GTF used to build the RSEM
reference:

```bash
rsemflow annotation gtf --i-study study/ --i-gtf Homo_sapiens.GRCh38.gtf.gz --o-annotation results/annotation.tsv
```

Without the GTF, the Bioconductor OrgDb packages map stable Ensembl IDs:

```bash
rsemflow annotation ensembl --i-study study/ --p-species human --o-annotation results/annotation.tsv
```

Join the annotation to any result table by `gene_id`.

## Everything that was written

```{r}
list.files("results", recursive = TRUE)
```
